Why do we Eval?\\[2mm]\Large A large-scale study of Eval usage in R

### 2020-12-17 paper/main.tex
main

To provide a picture of \eval usage in R, this section focuses on
library code and the \Corpus packages in the CRAN corpus. Data was
collected from \Nbruns runs of code extracted from the target packages
and their reverse dependencies. In total, \Allcalls invocations of
\eval and its variants were recorded. Runs that did not call \eval
were discarded.\footnote{The fact that a package contains call sites
to \eval does not mean every run will execute those call sites.} Out
of the \Staticatleastonecallsite packages that use \eval,
\Triggeredpkgs packages we recorded calling the function --- the
difference can be chalk down to incomplete code coverage and
occasional failures in our pipeline.

\begin{wrapfigure}{r}{5.2cm}
\begin{tabular}{|r@{\,}r@{\,}lr|r@{\,}r@{\,}lr|} \hline
\tt 0 &--& \tt 10      & \Bina  & \tt 1K &--&\tt 100K  & \Bine\\
\tt 11 &--& \tt 100    & \Binb  & \tt 100K &--&\tt 1M  & \Binf\\
\tt 101 &--& \tt 1K    & \Binc  & \tt 1M &--&\tt 10M   & \Bing\\
\tt 1K &--& \tt 10K    & \Bind  & \tt 10M &--& \tt 100M & \Binh\\\hline
\end{tabular}
\caption{Call frequency}\label{freq}
\end{wrapfigure}

Figure~\ref{freq} shows the number of \evals generated by each package
over all runs that invoked functions in that package. Out of the
\Corpus packages in the corpus, \Fewcalls packages have a low
frequency of \eval calls, with fewer than 100 of them. On the other
hand, \Manycalls packages call eval more than 1,000 times, with one
package (\Maxcallspack) calling it \Maxcalls times.

Figure~\ref{tab:variantseval} shows, per variant of \eval, the number
of call sites in the corpus (\emph{static}), the number of call sites
that were observed during tracing (\emph{exercised}), and the total
number of dynamic calls (\emph{invocations}). The data clearly shows
that the variants of \eval are rarely used, in particular \c{evalq}
and \c{local}.

\begin{figure}[h]\begin{tabular}{|r|r|r|r|r|}\hline
  &\eval & \c{evalq} & \c{eval} & \c{local}\\[-2mm]
           & & & \c{.parent} &\\\hline
Static sites &\Staticeval&\Staticevalq&\Staticevalparent&\Staticlocal \\
Exercised sites&\Triggeredeval&\Triggeredevalq&\Triggeredevalparent&\Triggeredlocal\\
Invocations&\EvalsRnd&\EvalqsRnd&\EparentsRnd&\LocalsRnd\\\hline
\end{tabular}\caption{Variants}\label{tab:variantseval}
\end{figure}

\begin{figure}[h] \centering
\begin{tabular}{|r@{\,}r@{\,}lr|r@{\,}r@{\,}lr|} \hline
\tt 0 &--& \tt 50    & \Runbina & \tt 501 &--&\tt 1000   & \Runbine\\
\tt 51 &--& \tt 100  & \Runbinb & \tt 1001 &--&\tt 1500  & \Runbinf\\
\tt 101 &--& \tt 250 & \Runbinc & \tt 1501 &--&\tt 2000  & \Runbing\\
\tt 251 &--& \tt 500 & \Runbind & \tt 2001 &--& \tt 3000 & \Runbinh\\\hline
\end{tabular}
\caption{Normalized invocations} \label{tab:callsitesnormalized}
\end{figure}

Figure~\ref{tab:callsitesnormalized} shows the number of invocation to
\eval and its variants per call sites per run exercising that call
site. If that number is high, it can correspond to a function with an
\eval call site that is used widely. If that number is very high, more
than 1000, we expect it to correspond to an \eval call site in a loop.
Most of the invocations are less than 50 per run per call site. When
zooming in, on Figure~\ref{fig:callsitesnormalized}, most call sites
have actually only one invocation per run.

\begin{figure}[h] \centering
    \includegraphics[width=.5\textwidth]{calls_per_run_per_call_site}
    \caption{Invocations of \eval per call sites normalized by the
      number of runs, for less than 50 normalized invocations. Binsize
      is 1} \label{fig:callsitesnormalized}
\end{figure}

>>>>

To provide a picture of \eval usage in R, this section focuses on
library code and the \Corpus packages in our CRAN corpus. For these
packages, all runnable code in CRAN was executed, runs that did not
exercise an \eval in the corpus were discarded. The data presented
here was collected from \Nbruns runs. In total, \Allcalls invocations
of \eval and its variants were recorded. Out of the
\Staticatleastonecallsite packages that use \eval, \Triggeredpkgs
packages were observed calling it --- the difference can be chalked
down to incomplete code coverage and occasional infrastructure
failures.

\begin{wrapfigure}{r}{5.2cm}
  \begin{tabular}{|r@{\,}r@{\,}l@{}r|r@{\,}r@{\,}l@{}r|} \hline
    \multicolumn{3}{|c}{\small\#calls} &\small \#pck
&     \multicolumn{3}{c}{\small\#calls} &\small\#pck \\\hline
\tt 0 &--& \tt 10      & \Bina  & \tt 1K &--&\tt 100K  & \Bine\\
\tt 11 &--& \tt 100    & \Binb  & \tt 100K &--&\tt 1M  & \Binf\\
\tt 101 &--& \tt 1K    & \Binc  & \tt 1M &--&\tt 10M   & \Bing\\
\tt 1K &--& \tt 10K    & \Bind  & \tt 10M &--& \tt 100M & \Binh\\\hline
\end{tabular}
\caption{Call frequency}\label{freq}
\end{wrapfigure}

Figure~\ref{freq} summarizes the frequency of calls to \eval over all
runs. The left column is the number of observed calls, the right
column is the number of packages falling in that range. There are
\Fewcalls packages with low \eval frequency, i.e. fewer than 100
calls; and a similar number of packages, \Manycalls to be precise, use
\eval more than 1,000 times. One package, \Maxcallspack, calls it
\Maxcalls times and thus accounts for over half of the observed
invocations of \eval.

\begin{wrapfigure}{r}{7cm}
  \begin{tabular}{|@{\,}r@{\,}|@{\,}rrrr@{\,}|}\hline
  &\eval & \c{evalq} & \c{eval} & \c{local}\\[-2mm]
           & & & \c{.parent} &\\\hline
\small Static sites &\Staticeval&\Staticevalq&\Staticevalparent&\Staticlocal \\
\small Exercised sites&\Triggeredeval&\Triggeredevalq&\Triggeredevalparent&\Triggeredlocal\\
\small Invocations&\EvalsRnd&\EvalqsRnd&\EparentsRnd&\LocalsRnd\\\hline
\end{tabular}\caption{Variants}\label{tab:variantseval}
\end{wrapfigure}

Figure~\ref{tab:variantseval} illustrates the use of variants of
\eval. For each of the four variants, the first rows shows the number
of call sites occurring in the corpus (\emph{static)}), the second row
is the number of call sites that were encountered during tracing
(\emph{exercised}), the last row is the number of calls
(\emph{invocations}). The vast majority of call sites, and calls, are
to \eval itself; \c{evalq} and \c{local} are barely used at all.

\begin{wrapfigure}{r}{5.2cm}
\begin{tabular}{|r@{\,}r@{\,}lr|r@{\,}r@{\,}lr|} \hline
\tt 0 &--& \tt 50    & \Runbina & \tt 501 &--&\tt 1000   & \Runbine\\
\tt 51 &--& \tt 100  & \Runbinb & \tt 1001 &--&\tt 1500  & \Runbinf\\
\tt 101 &--& \tt 250 & \Runbinc & \tt 1501 &--&\tt 2000  & \Runbing\\
\tt 251 &--& \tt 500 & \Runbind & \tt 2001 &--& \tt 3000 & \Runbinh\\\hline
\end{tabular}
\caption{Normalized invocations} \label{tab:callsitesnormalized}
\end{wrapfigure}

Figure~\ref{tab:callsitesnormalized} shows normalized invocation
counts. The column on the left is the average number of invocations of
\eval from a particular call site for any given run. The number on the
right is number of call sites that fall in that range. For instance,
there are \Runbinh call sites that are invoked on average over 2,001
times in any run they are exercised.

\begin{wrapfigure}{r}{5.2cm}
  \vspace{-8mm}
    \includegraphics[width=.4\textwidth]{calls_per_run_per_call_site}
    \caption{Normalized invocations
      (small)} \label{fig:callsitesnormalized}
    \vspace{-3mm}
\end{wrapfigure}

The majority of call sites, \Runbina, are exercised up to 50 times.
Larger numbers of calls suggest that \eval may be called within a loop
or from a recursive context -- this seems to be the exception.
Figure~\ref{fig:callsitesnormalized} zooms in on call sites with an
average of fewer than 51 invocation. The x-axis shows the number of
invocations and the y-axis is the number of call sites. Most call
sites are invoked once, about half as many are invoked twice. After
that the frequency quickly drops.
